---
title: "Introduction: a wave-by-wave sign flip"
---

```{r}
#| label: setup
#| include: false
source(here::here("_common.R"))
```

## The puzzle

A standard wave-by-wave logistic regression on Bangladesh DHS data spanning 2004–2022 (N = 33,732 children, 62 candidate predictors) returns a result that shouldn't be possible if the underlying causal mechanism is stable: the C-section coefficient *changes sign* across the eighteen-year observation window.

> **In 2004**, C-section is associated with **higher** child mortality.
> **In 2022**, C-section is associated with **lower** child mortality.

Two coefficients with opposite signs from the same descriptive model on overlapping populations is not noise. It is a structural finding — large enough to rule out sampling variance, structured enough to demand a causal investigation.

This book documents that investigation. A recursive bivariate probit (RBVP) estimator [@marra2011estimation; @greene2012econometric] is built and run; three formal identification diagnostics [@swanson2013iv; @terza2018twostage; @mcgovern2015robustness] are applied; a modern sensitivity layer [@vanderweele2017evalue; @cinelli2020sensemakr; @acerenza2023testable] tests the assumptions on which the headline estimate rests. The conclusion of the diagnostic chain is not "C-section reduces mortality." The conclusion is that **the identifying assumptions on which a causal interpretation depends are rejected by the data**.

## The substantive backdrop

The reason the sign flips, the book argues, is that the *data-generating process* underlying C-section selection changed dramatically between 2004 and 2022:

- 2004: C-section rate ≈ 4.5%. Surgery is *access-constrained*; women who reach a facility tend to be high-risk emergencies. Selection on severity dominates.
- 2022: C-section rate ≈ 44%, well past the WHO 15% threshold [@who2015csection; @sujon2025cesarean]. Surgery is *demand-driven*, increasingly elective, concentrated in private facilities and upper wealth quintiles.

This is a *regime transition*. Standard IV strategies that assume a fixed selection mechanism cannot identify a single coherent treatment effect across that span. Chapters 6–11 work out the formal consequences; chapter 19 extends the analysis to India and Pakistan; Chapter 20 names the methodological pivot the data implies.

## Sample composition

```{r}
#| label: tbl-samples
#| tbl-cap: "Sample composition by DHS wave, after Step-1 cleaning."
samples <- load_sample_chars()
samples |> book_table(digits = 0)
```

## Three reader paths through this book

For social scientists, the salient chapters are Part I (descriptive trajectory, regime transition, wealth gradient, mortality trends) and the modern-inference layer in Chapter 15 (predetermined Chow test, the regime-shift's structural visibility). For IV econometricians, Chapter 6 (the econometric framework), Chapter 9 (cross-instrument placebo), and Chapter 11 (Conley plausibly-exogenous bounds) carry the methodological argument. For statisticians, Chapter 7 (RBVP), Chapter 8 (2SRI Hausman test), Chapter 10 (copula sensitivity), and Chapter 14 (Acerenza testable inequalities) interrogate the bivariate probit's identifying assumptions.

The conclusion (Chapter 20) is written for all three audiences in turn.
