---
title: "IV validity: the cross-instrument placebo test"
---

```{r}
#| label: setup
#| include: false
source(here::here("_common.R"))
```

## Motivation

The exclusion restriction — that the instrument affects mortality *only through* the individual C-section decision — is the IV condition that cannot be tested directly. But it can be *falsified*. The cleanest falsification is the cross-instrument placebo test from Swanson & Hernán [@swanson2013iv]: if an instrument intended for treatment $D_1$ also predicts a different treatment $D_2$, then the instrument is not channeling its effect through $D_1$ alone — it's correlated with general healthcare access, social position, or some other community-level confound, and the exclusion restriction is violated.

This chapter runs that test for both treatments (C-section, ANC-4+) using both instruments (community-LOO C-section rate, community-LOO ANC rate). Two treatments × two instruments × four waves gives eight cells. A single failure is bad; eight is decisive.

## Data and setup

The diagnostics are implemented in `scripts/40_causal_main/44_iv_validity_tests.R`. The script also runs first-stage F statistics, balance regressions against observable confounders, and the reduced-form regression of the outcome on the IV. The cross-instrument placebo is the single most direct test of exclusion.

## Results

```{r}
#| label: tbl-placebo
#| tbl-cap: "Cross-instrument placebo: every IV → wrong-treatment regression. A pass requires the IV to predict its own treatment but NOT the other. Bold-marked fails reject the exclusion restriction."
plac <- load_iv_placebo()
make_placebo_table(plac)
```

## Interpretation

Eight cells. Eight failures. Every one at p < 10⁻⁷.

The community-LOO C-section instrument strongly predicts ANC-4+ use in every wave; the community-LOO ANC instrument strongly predicts C-section in every wave. Both instruments are reflecting *general* healthcare access at the cluster level, not channeling their effect through their own treatment alone.

This is the most direct evidence in the book that the exclusion restriction is violated. It is not a power problem (the p-values are at scientific-notation strength). It is a structural one: at the community level, the two instruments and the two treatments are too entangled to be separated. Angrist's critique of leave-one-out community-mean instruments [@angrist2014peer] applies; Betz, Cook & Hollenbach's critique of spatial aggregation [@betz2018spatial] applies.

Whatever the 2022 RBVP coefficient is estimating, it is not the causal effect of C-section on mortality holding all else equal. It is a *bundled* effect of general healthcare-access expansion in a regime-shifting environment.

## Code

```{r}
#| file: scripts/40_causal_main/44_iv_validity_tests.R
#| eval: false
#| code-fold: true
#| code-summary: "scripts/40_causal_main/44_iv_validity_tests.R — full IV-validity diagnostic battery"
```
