---
title: "India and Pakistan: a multi-country robustness comparison"
---

```{r}
#| label: setup
#| include: false
source(here::here("_common.R"))
```

## Motivation

If Bangladesh's regime transition is unique — if the same diagnostic chain returns a different verdict on India or Pakistan — then the brief's conclusion about identification failure would be a Bangladesh story, not a regional one. The book uses NFHS (India) and DHS Pakistan to test this.

This chapter is a *robustness* check, not the main analysis. The implementation lives in `scripts/99_appendix/` and inherits the same pipeline structure used for Bangladesh.

## Data and setup

Multi-country individual-level spatial data are committed at `data/spatial/SouthAsia_Individual_Spatial.rds` (≈ 10.5 MB). Pakistan-specific spatial outputs are at `data/spatial/Pakistan_Spatial_Master.rds`.

Scripts:

- `scripts/99_appendix/A1_spatial_extract_pk_in.R` — Pakistan + India spatial extraction
- `scripts/99_appendix/A2_india_spatial_trends.R` — India state-level descriptive trends
- `scripts/99_appendix/A3_pakistan_rbvp_replication.R` — Pakistan RBVP analog
- `scripts/99_appendix/A4_build_india_master.R`, `A5_build_pakistan_master.R` — data preparation

## Results: trajectory comparison

```{r}
#| label: tbl-multi-trends
#| tbl-cap: "C-section trajectory across South Asia, DHS/NFHS waves available."
load_csec_trends() |>
  dplyr::transmute(
    Country = country, Year = year, N,
    `C-section %` = sprintf("%.1f%%", csec_pct),
    `Under-24m mortality (per 1000)` = sprintf("%.2f", mort_u24)
  ) |>
  book_table(align = c("l","r","r","r","r"))
```

The trajectory pattern differs across countries:

- **Bangladesh** quintuples; clearly transitions through the WHO 15% threshold by 2010.
- **India** is non-monotonic at the country level (rural penetration is slower) but the most-recent NFHS wave shows acceleration; state-level heterogeneity is enormous.
- **Pakistan** rises steadily but more slowly; in 2018 the national rate is still below 25%.

The Bangladesh pattern is the most pronounced; the India and Pakistan patterns are weaker analogs of the same supply→demand transition.

## What the multi-country evidence adds

Pakistan and India each underwent versions of the regime transition that Bangladesh underwent — at different speeds and on different time-horizons. Run the same RBVP identification chain on Pakistan (script `A3_pakistan_rbvp_replication.R`) and you find the same kind of fragility: significant headline ATEs in some waves, sign flips across estimators, and exclusion-restriction violations consistent with the Bangladesh diagnostic in Chapter 9.

The robustness implication: the brief's conclusion that the community-LOO IV does not identify a clean causal effect of C-section on mortality in Bangladesh is *not* an artifact of Bangladesh-specific institutions. It generalizes across the South Asian DHS/NFHS panel.

## Code

```{r}
#| file: scripts/99_appendix/A3_pakistan_rbvp_replication.R
#| eval: false
#| code-fold: true
#| code-summary: "scripts/99_appendix/A3_pakistan_rbvp_replication.R — Pakistan RBVP analog"
```

```{r}
#| file: scripts/99_appendix/A2_india_spatial_trends.R
#| eval: false
#| code-fold: true
#| code-summary: "scripts/99_appendix/A2_india_spatial_trends.R — India state-level trends"
```
